---
title: "Ispit 11.06.2024"
output: html_notebook
---

# Exercise 9

ะะฐ ะฟัะธะผะตัะพะบ ะพะด ๐1 = 10 ะฟัะพะธะทะฒะพะดะธ ะพะด ะตะดะฝะฐ ะผะฐัะธะฝะฐ ะต ััะฒัะดะตะฝะพ ะดะตะบะฐ ะฟัะพัะตัะฝะฐัะฐ ะดะพะปะถะธะฝะฐ
ะฝะฐ ะฟัะพะธะทะฒะพะดะพั ะต 70 ะผะผ ะฐ ะฟัะตัะผะตัะฐะฝะฐัะฐ ะดะธัะฟะตัะทะธัะฐ ะทะฐ ะธััะธะพั ะฟัะธะผะตัะพะบ ะต $๐๐^2_1$ = 5.7๐๐. ะะพ ะฝะตะบะพะต ะฒัะตะผะต ะต ะฝะฐะฟัะฐะฒะตะฝ ัะตะผะพะฝั ะฝะฐ ะผะฐัะธะฝะฐัะฐ ะธ ะฟะพะฒัะพัะฝะพ ะต ะทะตะผะตะฝ ะฟัะธะผะตัะพะบ ัะพ ๐2 = 15 ะฟัะพะธะทะฒะพะดะธ ัะพ ััะตะดะฝะฐ ะฒัะตะดะฝะพัั 80 ะธ ะดะธัะฟะตัะทะธัะฐัะฐ $๐๐^2_2$ = 9.6๐๐.

ะ) ะะฐะปะธ ะผะพะถะต ัะพ ะฝะธะฒะพ ะฝะฐ ะทะฝะฐัะฐัะฝะพัั ฮฑ=0.05 ะดะฐ ัะผะตัะฐะผะต ะดะตะบะฐ ะดะพัะปะพ ะดะพ ะทะฝะฐัะฐัะฝะพ
ะทะณะพะปะตะผัะฒะฐัะต ะฝะฐ ะดะธัะฟะตัะทะธัะฐัะฐ ะฝะฐ ะฟัะพะธะทะฒะพะดะพั?

$$
H_0: \sigma^2_X = \sigma^2_Y
$$

$$
H_a: \sigma^2_X \neq \sigma^2_Y
$$

```{r}
n1 <- 10
av_1 <- 70
sn1 <- 5.7
n2 <- 15
av_2 <- 80
sn2 <- 9.6

fisher_test <- sn1/sn2
fisher_test

alpha <- 0.05
bound <- 1 / qf(1 - alpha, df1 = n2 - 1, df2 = n1 - 1)
fisher_test > +Inf && fisher_test < bound
# since the test is not in the critical domain, H_0 is rejected
```

ะ) ะะบะพ ะฟัะตัะฟะพััะฐะฒะธะผะต ะดะตะบะฐ ะดะธัะฟะตัะทะธัะฐัะฐ ะฝะฐ ะดะพะปะถะธะฝะฐัะฐ ะฝะฐ ะฟัะพะธะทะฒะพะดะพั ะฝะต ัะต ะฟัะพะผะตะฝะธะปะฐ, ัะพ
ะฝะธะฒะพ ะฝะฐ ะทะฝะฐัะฐัะฝะพัั ะพะด 0.05 ะดะฐ ัะต ัะตััะธัะฐ ะดะฐะปะธ ะธะผะฐ ะฟัะพะผะตะฝะฐ ะฝะฐ ะพัะตะบัะฒะฐะฝะฐัะฐ ะดะพะปะถะธะฝะฐ ะฝะฐ
ะฟัะพะธะทะฒะพะดะพั ะฟะพัะปะต ะฝะฐะฟัะฐะฒะตะฝะธะพั ัะตะผะพะฝั ะฝะฐ ะผะฐัะธะฝะฐัะฐ ะฒะพ ะพะดะฝะพั ะฝะฐ ะพัะตะบัะฒะฐะฝะฐัะฐ ะดะธะผะตะฝะทะธัะฐ
ะฟัะตะด ัะตะผะพะฝัะพั.

$$
H_0: \mu_X = \mu_Y,
$$

$$
H_a: \mu_X \neq \mu_Y
$$

```{r}
alpha <- 0.05
sp_squared <- ( (n1 - 1)*sn1 + (n2-1)*sn2 ) /(n1 + n2 -2)
sp_squared
sp <- sqrt(sp_squared)
t_0 = (av_1 - av_2) / (sp * sqrt((1/n1) + (1/n2)))
t_0
bound <- qt(alpha / 2, df=n1+n2-2, lower.tail = FALSE)
bound
in_lower <- t_0 > -Inf && t_0 < -bound
in_upper <- t_0 < -Inf && t_0 > bound
domain <- c(in_lower, in_upper)
domain
# since t is in the critical domain H_0 is rejected
```
